Comment: Reasoning about Interference Between Units

### 2014-10-16 paper/introduction.tex
Starting writing

 \citet{bowers2013sutva} showed that one could use Fisher's
 randomization-based hypothesis testing framework to assess models of
 treatment propagation and spillover across social networks. Working in a
 frequentist framework, they showed that one could make statements such as:
 "Our design provides strong evidence against the idea that the treatment
 spilled over by more than 5 points given this model of treatment
 propagation." This research note shows that the test statistic that they used
 --- the Kolmogorov-Smirnov test statistic  \citep[\S
 5.4]{MylesHollander1999a} --- could be improved substantially. We replicate
 their analyses to demonstrate that the sum of squared residuals from a linear
 model in which aspects of the network itself play a role.

 They also showed that the strength of evidence against the
 specific hypotheses implied by a given model varied with different features
 of the design as well as the extent to which the true process by which
 treatment given to one unit changed the outcomes of another unit diverged
 from the model. Throughout their paper they used the  Kolmogorov-Smirnov (KS)
 test statistic so that their tests would be sensitive differences in the
 treatment and control distributions implied by different hypotheses and not
 merely sensitive to differences in one aspect of those distributions (such as
 the differences in the mean).\footnote{If the empirical cumulative distribution function (ECDF) of
   the treated units is $F_1$ and the ECDF of the control units is $F_0$ then
   the KS test statistic is
   $\T(\yu,\bz) = \underset{i = 1,\ldots,n}{\text{max}}\left[F_1(y_{i,\bzero})
     - F_0(y_{i,\bzero})\right]$, where $F(x)=(1/n)\sum_{i=1}^n I(x_i \le x)$
   records the proportion of the distribution of $x$ at or below $x_i$
   \citep[\S 5.4]{MylesHollander1999a}.}

 provides both clarity for
 non-parametric statistical inference  also implied great flexibility in regard to the scientific questions
that one might engage: for example, one could use the sharp null itself to
detect spillover  among units \citep{aronow-general} or to characterize the
benefits from treatment on treated units even if  and sharp hypotheses
about effects (in contrast to hypotheses about no effects) have been developed
to assess models of interference \citep{bowers2013sutva}, [ include Rosenbaum
2007, Bowers and Hansen 2009, and a bunch of other stuff by Rosenbaum ].

Fisher's sharp null hypothesis of no effects combined with his
randomization-based hypothesis testing framework provides a very clear and
clean vision of what it means to test a hypothesis \citep[Chap
2]{fisher:1935} and \citep{rosenbaum2010design}.
One problem that has become apparent, as scholars have extended Fisher's basic
ideas, is that as the models of effects have become more complex, the power of
the testing procedure itself has become more important. A simple test of the
sharp null can be executed with power that varies as a function of the design
of the study (proportion treated, blocking structure, etc..), characteristics
of the outcome, and the way that a test statistic summarizes the outcome. Yet,
when the models are complex,  \citet{bowers2013sutva} showed that sometimes a
test will have (1) no power to address the model at all (a function, in part
of the model itself --- for example ...) such that all hypothesized parameters
would receive the same implausibility assessment; (2) or might reject all such
hypothesized parameters. Thus, although in theory one may assess sharp
hypotheses about multiparameter settings, in practice one may not learn much
from such tests --- or, in the absence of simulation studies assessing the
model and test statistic, one may be mislead.

In this paper we use a very simple causal model and a simple test statistic to
first show how the combination of model and test statistic is a crucial choice
--- and how more attention ought to be paid to both the model and the test
statistic.

Then we suggest one test statistic that we think should have good properties
when the number of parameters in a model is low (i.e. $k << n$). And we should
this to be the case --- that the robust F-stat from a linear model is a better
way to go than the KS-test and certainly better than any single measure of
distributional difference.

Finally, we highlight the similarities between testing the sharp hypothesis
and the weak hypothesis --- which in turn allows even more flexibility and
perhaps better performing testing procedures.

Then we suggest a few alternative avenues to pursue.

Some models are well powered for some designs and not other designs.

Using same simulation study using same data from PA study. Better than KS
test.

The F-test is model consistent. By design it controls Type I error, but if the
model is correct, it has some optimality criterion. If you had control
variables, for example, then it would be consistent against mean differences.

Under the model, model consistent.

>>>>

\citet{bowers2013sutva} showed that one could use Fisher's
randomization-based hypothesis testing framework to assess counterfactual
causal models of treatment propagation and spillover across social networks.
That is, they showed that one could make
statements such as: ``Our design provides strong evidence against the idea
that the treatment spilled over by more than 5 points given this model of
treatment propagation.'' This research note shows that the test statistic that
they used --- the Kolmogorov-Smirnov test statistic  \citep[\S
5.4]{MylesHollander1999a} --- could be improved substantially. We demonstrate that the sum of squared residuals from a linear
model in which aspects of the network play a role is a preferable test
statistic for the model and network and design analyzed by
\citet{bowers2013sutva} (henceforth BFP). The point of this short article is to improve the
use of the ``reasoning about interference'' approach and also to highlight
the methodological and statistical questions raised by that article. We do
not aim for general answers to those questions here and now, but we do hope
that this piece stimulates further, and  deeper, work on test statistics and
sharp hypotheses.

\section{Background}

We can define a counterfactual causal effect for a person $i$ as a difference
between the outcome, $y_i$, person $i$ would have expressed
after a set of treatments, collected in the vector $\bz$, had been assigned to some subset of a social network,
$y_{i,\bz}$ compared to the outcome that person $i$ would have expressed
under a different treatment vector. If the experiment had an effect on person
$i$, then her outcome would differ under different realizations of the
experimental treatment $y_{i,\bz} \ne y_{i,\bz'}$. The fundamental problem of
causal inference (a phrase famously coined in \cite{holland:1986a}) reminds
us that we can never see both states of the world: we only observe the
outcome from person $i$ under one treatment assignment, either $\bz$ or
$\bz'$ not both.

R.A. Fisher's approach to design-based statistical inference  \citep[Chap
2]{fisher:1935} begins with the idea that we do not learn directly about the
unobserved counterfactuals. Rather, he suggests, we can learn about models of
unobserved counterfactuals --- we can learn how much information we have to
dispel certain claims about those counterfactuals as implausible. This
conceptual move --- sidestepping the fundamental problem of causal inference
via learning about claims made by scientists --- drives hypothesis testing in
general.  BFP build on this insight as developed and
linked to counterfactual causal inference by Paul Rosenbaum
\citep{rosenbaum2010design}. They also say: we do not know how the
counterfactual conditions might have turned out, they are unobserved, but we
can make claims about how treatment might have come to have an effect and we
can assess those claims. For example, the claim used by
BFP is a model in which the effects of treatment
propagating from one person to another die off as the network distance between
nodes increases.\footnote{See the paper itself for details of the example
  model.} They further showed that the strength of evidence against the
specific hypotheses implied by a given model varied with different features of
the design as well as the extent to which the true process by which treatment
given to one unit changed the outcomes of another unit diverged from the
model. Throughout their paper they used the Kolmogorov-Smirnov (KS) test
statistic so that their tests would be sensitive differences in the treatment
and control distributions implied by different hypotheses and not merely
sensitive to differences in one aspect of those distributions (such as the
differences in the mean).\footnote{If the empirical cumulative distribution
  function (ECDF) of the treated units is $F_1$ and the ECDF of the control
  units is $F_0$ then the KS test statistic is $\T(\yu,\bz) = \underset{i =
    1,\ldots,n}{\text{max}}\left[F_1(y_{i,\bzero}) -
    F_0(y_{i,\bzero})\right]$, where $F(x)=(1/n)\sum_{i=1}^n I(x_i \le x)$
  records the proportion of the distribution of $x$ at or below $x_i$
  \citep[\S 5.4]{MylesHollander1999a}.}

Notice that the test statistic choice matters: the engine of statistical
inference involves summarizing information against the hypothesized claim, yet
different ways to summarize information might be more or less sensitive to
substantively meaningful differences. A simple test of the
sharp null can be executed with power that varies as a function of the design
of the study (proportion treated, blocking structure, etc..), characteristics
of the outcome, and the way that a test statistic summarizes the outcome. Yet,
when the models are complex,  BFP showed that sometimes a
test will have (1) no power to address the model at all such that all hypothesized parameters
would receive the same implausibility assessment; (2) or might reject all such
hypothesized parameters. Thus, although in theory one may assess sharp
hypotheses about multiparameter settings, in practice one may not learn much
from such tests --- or, in the absence of simulation studies assessing the
model and test statistic, one may be mislead.

BFP recommends choosing ``effect increasing" test statistics ``that will be
small when the treated and control distributions in the adjusted data ... are
similar, and large when the distributions diverge." However, test statistics
that only compare distributional features of hypothesized potential outcomes
between individuals assigned to treatment and those assigned to control fail
to fully exploit the data at hand. The problem is that these test statistics
are  not necessarily increasing in spillover effects.

I demonstrate two results, each showing how effect increasing test statistics,
as defined by BFP, may provide little to no information about the spillover
parameter $\tau_2$.  Let us first consider the case when the test statistic is
the absolute mean difference between hypothesized uniformity trial outcomes
for treated and control individuals: $\T(\widetilde y_{ij}(\0,\Z), \Z') =
\left\vert \mu_1 ( \widetilde y_{ij}(\0,\Z), \Z' ) - \mu_0 ( \widetilde
  y_{ij}(\0,\Z), \Z' ) \right\vert.  $ Fix a realization of $\Z = \z$, and
further assume noncollinearity: $ \z \notin \{\z^I, (1-\z^I)\}$. We can see
that, for any hypothesized $\widetilde\tau_2$, there always exists a
$\widetilde\tau_1$ such that the $p$-value is 1. Expanding the test statistic,
$\T(\widetilde y_{ij}(\0,\z), \z) = \left\vert \mu_1 (y_{ij}(\z) -
  \widetilde\tau_1 - \tau_2 z_{i(-j)}, \z) - \mu_0(\widetilde y_{ij}(\z) -
  \widetilde\tau_2 z_{i(-j)} , \z) \right\vert = \left\vert \mu_1 (y_{ij}(\z)
  - \widetilde\tau_2 z_{i(-j)}, \z)- \mu_0 (\widetilde y_{ij}(\z) -
  \widetilde\tau_2 z_{i(-j)}, \z)- \widetilde\tau_1 \right\vert.  $ Thus,
regardless of the value of $\widetilde\tau_2$, we can always find a
$\widetilde\tau_1$ that sets $\T(\widetilde y_{ij}(\0,\z), \z) = 0$. Given
that $\T(\widetilde y_{ij}(\0,\z), \Z')$ can only take values in
$\mathbb{R}^+$, then $p=1$, as any other possible randomization $\Z'$ will
yield a value of $\T(\widetilde y_{ij}(\0,\z), \Z') \geq 0$. It follows that,
given this test, we cannot rule out any value of $\tau_2$ as being
implausible.\footnote{In fact, this result does not depend on example in the
  setting, and holds for any randomization scheme coupled with a linear causal
  model of the form \eqref{causal}.}

Second, I show that, given the running example, a test based on any effect
increasing statistic is inconsistent against a radically incorrect hypothesis
about $\tau_2$. Let us assume that $\tau_1$ is properly hypothesized at its
true value, and that the hypothesized value of $\tau_2$ ($\widetilde \tau_2$)
is radically incorrect: $\vert \widetilde\tau_2 - \tau_2 \vert \gg
\max(y_{ij}(\0)) - \min(y_{ij}(\0))$. Denote $\beta \equiv \tau_2 -
\widetilde\tau_2$, and assume $\vert \beta \vert < \infty$.  Under the
assumption that $\tau_2 = \widetilde \tau_2$: $\widetilde y_{ij}(\0,\Z) =
y_{ij}(\0) + \beta Z_{i(-j)}$.  Assume a sequence of nested finite populations
$\{U_k\} = U_1, U_2,...$, growing in $k$ households, so that $U_k$ consists of
$N_k = 2k$ individuals, each with an associated random assignment $\Z_k$
\citep{isaki1982survey}. (For notational parsimony, I drop explicit dependence
on the subscript $k$.)  Given an incorrect hypothesis, a necessary, but not
sufficient, condition for a test's consistency is that $\lim_{k \rightarrow
\infty} \Pr(p > \alpha) < \epsilon$, for any $\alpha,\epsilon > 0$.

Let an effect increasing test statistic $\T(\widetilde y_{ij}(\0,\Z),\Z')$ be
some measure of the absolute divergence in empirical distributions of
$\widetilde y_{ij}(\0,\Z)$ for treated and control individuals under the
treatment vector $\Z'$.  Since distributional differences in $y_{ij}(\0)$ are
small relative to shifts induced by differences in proportions of treated
housemates,  $\T(\widetilde y_{ij}(\0,\Z),\Z')$ will be increasing in $
\left\vert \mu_1(Z_{i(-j)},\Z') -   \mu_0(Z_{i(-j)},\Z') \right\vert$. , For
a realization $\Z = \z$, the hypothesized randomization distribution of
$\mu_1(z_{i(-j)},\Z') - \mu_0(z_{i(-j)},\Z')$ is asymptotically normal with
mean $0$ and variance $V_{x(\z)} = 4 x(\z) \left[1-x(\z)\right] / N$, where
$x(\z) \equiv \sum_{ij\in U}z _{ij} /N$, if $0 < x(\z) <  1$.  (by the
Lindeberg CLT, Taylor linearization and Slutsky's theorem).  Thus the
$p$-value associated with randomization $\z$, $p = 2
\Phi(-|\mu_1(z_{i(-j)},\z) - \mu_0(z_{i(-j)},\z)|/V_{x(\z)}^{1/2})$. It
follows that $|\mu_1(z_{i(-j)},\z) - \mu_0(z_{i(-j)},\z)\vert <
-\Phi^{-1}(\alpha/2) V_{x(\z)}^{1/2} \Leftrightarrow p > \alpha$.  By similar
calculations at the household level, the (actual, not hypothesized)
randomization distribution of $\mu_1(Z_{i(-j)},\Z) - \mu_0(Z_{i(-j)},\Z)$ is
asymptotically mean $0$ with variance $2 V_{1/2} =  2/N$.\footnote{This result
  follows from reframing the problem as randomly assigning households into 4
  conditions \{00,01,10,11\} with equal probability, then linearizing
  $\mu_1(Z_{i(-j)},\Z) - \mu_0(Z_{i(-j)},\Z)$.} By the LLN ,  $\plim_{k
\rightarrow \infty} x = 1/2$, and so, by the and continuous mapping theorem,
$\lim_{k \rightarrow \infty} \Pr(\vert NV_{x(\Z)} - NV_{1/2}\vert >
\epsilon_3) = 0$.  Then, by Chebyshev's inequality, $\lim_{k \rightarrow
\infty}  \Pr(p > \alpha) > 1- 2\left[\Phi^{-1}(\alpha/2)\right]^{-2}$,
$\epsilon = 1- 2\left[\Phi^{-1}(\alpha/2)\right]^{-2}$ proportion of possible
$\Z \in \Omega$ have $p > \alpha$,
for any $0 < \alpha < 2\Phi(-\sqrt{2})$.  Tests based on effect increasing
statistics are therefore not generally consistent.

\subsection*{Alternative test statistics}

There are usually better ways to evaluate the fit of a model than comparing
two marginal distributions. One simple test statistic would be the
sum-of-squared-residuals (SSR) from a least squares regression of $\widetilde
y_{ij}(\0,\Z)$ on $Z_{ij}$ and $Z_{i(-j)}$. Since the model in \eqref{causal}
is linear, $\T(\widetilde y_{ij}(\0,\Z), \Z) = \sum_{ij \in U} \left[
  \widetilde \Y(\0) - (\X^T\X)^{-1}\X^T \widetilde \Y(\0) \right]^2$, where $$
\widetilde \Y(\0) \equiv \left( \begin{array}{c} \widetilde y_{11}(0) \\
    \vdots \\ y_{(N/2)(-1)}(0) \end{array} \right), \X \equiv  \left(
  \begin{array}{ccc} 1 & Z_{11} & Z_{1(-1)} \\ \vdots & \vdots & \vdots  \\ 1
    & Z_{(N/2)(-1)} & Z_{(N/2)1} \end{array} \right) = ({\bf 1} \; \Z \;
\Z^I).  $$ Assume now that either the hypothesized $\tau_1$ or $\tau_2$ is
radically incorrect: i.e., $\vert \widetilde\tau_1 - \tau_1 \vert$ or $\vert
\widetilde\tau_2 - \tau_2 \vert \gg \max(y_{ij}(\0)) - \min(y_{ij}(\0))$.
Fixing $\Z = \z$, for any randomization $\Z' \notin \{\z, (1-\z), \z^I,
(1-\z^I), \0, \one \}$, $\T(\widetilde y_{ij}(\0,\z),\Z') > \T(\widetilde
y_{ij}(\0,\z),\z)$. Assuming noncollinearity,
%$\Z \notin \{\Z^I, (1-\Z^I)\}$, , (0,...,0), (1,...,1)\}$)
$p \leq 6  \left( 2^{-N} \right)$, rejecting the hypothesis with $\alpha =
0.05$ for any $N \geq 8$. %Consistency follows straightforwardly.

In the example above, use of SSR as a test statistic has maximal power to
reject radically incorrect hypotheses, whereas the test statistics proposed by
BFP are inconsistent. But SSR from an ordinary least squares regression is not
always appropriate: for example, when the probability of exposure to spillover
is heterogeneous across individuals, we may wish to apply inverse probability
weights so as to ensure representative samples of potential outcomes. This
suggests a conjecture: that the $SSR$ from an {\it
  inverse-probability-weighted} least squares regression is more generally a
sensible test statistic for models that include interference.  (The nature of
these weights will be discussed when discussing alternative modes of
inference.) Additionally, when nonlinear deviations from model predictions are
of concern, a weighted variant of the Brownian distance covariance
\cite{szekely2009brownian} or other {\it E}-statistic may be more sensible
than SSR.

\begin{figure}[h!]
\centering
\includegraphics[width=.9\textwidth]{twoDplots.pdf}
\caption{Proportion of $p$-values less than .05 for tests of joint hypotheses
about $\tau$ and $\beta$. Darker values mean rare rejection. White means
rejection always. Truth is shown at the intersection of the straight
lines  $\tau=.5, \beta=2$. Each panel shows a different test statistic. The
SSR Tests refer to eq, the KS tests refer to eq.  }\label{fig:ks}
\end{figure}

Test statistic selection is not trivial and BFP's contribution has effectively
opened a new line of research. Via counterexample, we can see that effect
increasing test statistics are not always appropriate for assessing hypotheses
about interference. BFP is correct in the assertion that, regardless of the
choice of test statistic selection, a set of implausible hypotheses is
identified by the procedure. But we should not be led to believe that, for any
given test statistic, that some hypotheses are more plausible than others.
Such inferences -- comparing hypotheses -- may depend on the test statistic
used, and not necessarily reflect the plausibility of the model at hand. This
point will gain importance as we now consider the limitations of sharp null
hypotheses.
